Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-09-06 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
checking from section 6. Part 1.

An early unpublished draft of Section \ref{newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek, on the other hand were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{AndersonHajekControversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
relesed, respectively, in Isabellle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of Gödel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of Gödel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled it to be
verified with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a so far failed attempt to formalize Avicenna's
argument in PVS as well.

@@

@@

					\AXC{$ $} \RightLabel{5}
					\UIC{$G(\gamma)$} \dottedLine
					\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
					\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$} \RightLabel{$\imp_E$}
				\BIC{$\neg \rho(\gamma)$} \RightLabel{$\neg_E$}      
	   \BIC{$\bot$} \RightLabel{$\neg_I^1$}
	   \UIC{$\neg\neg P(\rho)$} \RightLabel{$\neg\neg_E$}
	   \UIC{$[\subproof_4] \qquad P(\rho)$}
	   \end{prooftree}
\end{scriptsize}

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$G(\gamma)$}

 \AXC{$\subproof_4$}\dashedLine
 \UIC{$P(\rho)$}
	      \AXC{Axiom 4} \dashedLine
        \UIC{$\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]$} \RightLabel{$\all_E$}
	   	  \UIC{$P(\rho) \imp \nec P(\rho)$} \RightLabel{$\imp_E$}
		 \BIC{$\nec P(\rho)\scott$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\omega: P(\rho)$}
					\AXC{$ $} \RightLabel{2}
					\UIC{$\omega: G(\delta) $} \dottedLine
					\UIC{$\omega: \all \varphi. P(\varphi) \imp \varphi(\delta)$} \RightLabel{$\all_E$} 
					\UIC{$\omega: P(\rho) \imp \rho(\delta)$} \RightLabel{$\imp_E$}
				\BIC{$\omega: \rho(\delta) $} \RightLabel{$\imp_I^2$}
				\UIC{$\omega: G(\delta) \imp \rho(\delta) $} \RightLabel{$\all_I$}
				\UIC{$\omega: \all y. G(y) \imp \rho(y) $} \RightLabel{$\imp_I^3$}
				\UIC{$\omega: P(\rho) \imp (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\nec_I$}
				\UIC{$\nec (P(\rho) \imp (\all y. G(y) \imp \rho(y) ))$} \RightLabel{$\imp_E$}
						\AXC{Axiom K} \RightLabel{$\imp_E$}
					\BIC{$\nec P(\rho) \imp \nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\imp_E$}
			\BIC{$\nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\imp_I^4$}
			\UIC{$\rho(\gamma) \imp \nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\all_I$}
			\UIC{$\all \varphi. \varphi(\gamma) \imp \nec (\all y. G(y) \imp \varphi(y) )$} \RightLabel{$\wedge_I$}
		\BIC{$ G(\gamma) \wedge (\all \varphi. \varphi(\gamma) \imp \nec (\all y. G(y) \imp \varphi(y) ))$} \dottedLine
		\UIC{$\ess{G}{\gamma}$} \RightLabel{$ \imp_I^5$}
		\UIC{$G(\gamma) \imp \ess{G}{\gamma}$} \RightLabel{$\all_I$}
		\UIC{$\all x. G(x) \imp \ess{G}{x}$}
\end{prooftree}
\end{scriptsize}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. [P(\neg \varphi) \biimp \neg P(\varphi)]$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\wedge_E$}
\UIC{$\neg P(\rho) \imp P(\lambda x. \neg \rho(x))$}\RightLabel{$\imp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$P(\rho)$}
\AXC{Axiom \ref{N:A4}} \dashedLine
\UIC{$\all \varphi.[P(\varphi) \imp \nec\; P(\varphi)]$}\RightLabel{$\forall_E$}
\UIC{$P(\rho) \imp \nec\; P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$}\RightLabel{$\imp_I^3$}
\UIC{$[\subproof_7] \qquad \rho(\gamma) \imp \nec\; P(\rho)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec P(\rho)$} \RightLabel{$\nec_E$}
\UIC{$\omega: P(\rho)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\omega: G(y)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\omega: \forall \varphi.(P(\varphi) \imp \varphi(y))$}\RightLabel{$\forall_E$}
\UIC{$\omega: P(\rho) \imp \rho(y)$}\RightLabel{$\imp_E$}
\BIC{$\omega: \rho(y)$} \RightLabel{$\imp_I^4$}
\UIC{$\omega: G(y) \imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega: \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y.(G(y) \imp \rho(y))$} \RightLabel{$\imp_I^5$}
\UIC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$} \RightLabel{$\imp_E$}
\AXC{$ $} \RightLabel{6}
\UIC{$\rho(\gamma)$} \RightLabel{}
\AXC{$\subproof_7$} \dashedLine
\UIC{$\rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \RightLabel{$\wedge_I$}
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{N:D2}}
\UIC{$[\subproof_8] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{small}

@@

\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$} \RightLabel{}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec\rho(\gamma)$} \RightLabel{}
\BIC{$P(\rho)$} \RightLabel{$\imp_E$}

@@

\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$\nec P(\rho)$} \RightLabel{$\nec_E$}
\UIC{$\omega: P(\rho)$}
\AXC{$ $} \RightLabel{2}
\UIC{$\omega: G(y)$}  \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\omega: \forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$\omega: P(\rho) \biimp \nec \rho(y)$}\RightLabel{$\biimp_E$}
\BIC{$\omega: \nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega: \rho(y)$} \RightLabel{$\imp_I^2$}
\UIC{$\omega: G(y) \imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega: \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y.(G(y) \imp \rho(y))$} \RightLabel{$\imp_I^3$}
\UIC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$} \RightLabel{$\imp_E$}

\AXC{$\subproof_7$} \dashedLine
\UIC{$\nec P(\rho)$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^4$}
\UIC{$[\subproof_8] \qquad \nec\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\end{prooftree}

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} 
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_6$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. [P(\varphi)\biimp \nec \varphi(\gamma)]$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_9] \quad \nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

\begin{tiny}
\begin{prooftree}
\AXC{$\subproof_8$ and $\subproof_9$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)]) $}
\AXC{$ \subproof_6 $} \dashedLine
\UIC{$G(\gamma)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\biimp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{tiny}
\end{proof}

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{2}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^2$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec [G(\gamma)\imp \ex x.G(x)]$} \doubleLine  \RightLabel{(using Axiom K)}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
\end{scriptsize}

>>>>

An early unpublished draft of Section \ref{newproof} of this paper,
prepared by Woltzenlogel Paleo in March 2013, served as the starting
point for two distinct but complementary research endeavours. Siders
and Woltzenlogel Paleo transformed that early draft into this paper,
aiming for human-readable and yet formal natural deduction proofs.
Benzm\"uller and Woltzenlogel Paleo, on the other hand, focused on the
verification of variants of G\"odel's ontological arguments using
higher-order automated reasoning tools, which quickly led to many
publications. The main automatically obtained results about G\"odel's
and Scott's proofs appeared in the European Conference on Artificial
Intelligence \citep{ECAI2014} and, two years later, a detailed
analysis of an inconsistency found in G\"odel's axioms was published
in the International Joint Conference on Artificial Intelligence
\citep{IJCAI2016}. The automatic verification of variants by Anderson,
Bjordal and H\'ajek, on the other hand were presented in the First
World Congress on Logic and Religion
\citep{WorldCongressLogicAndReligion} and published in
\citep{AndersonHajekControversy} and results related to the modal
collapse were reported in the 4th Word Congress on the Square of
Opposition \citep{SquareOfOpposition} and later published in
\citep{ModalCollapse}. More detailed reconstructions of Scott's
variant within the interactive proof assistants Isabelle and Coq were
relesed, respectively, in Isabellle's Archive of Formal Proofs
\citep{IsabelleGod} and in the Symposium of Computer Science in Russia
\citep{CSRModalLogicInCoq}. The popularity of the topic led to invited
talks and corresponding invited abstracts in the proceedings of
workshops \citep{WADT,WADTPreProceedings,IWIL} and summer schools
\citep{SpiritualQuest,ReasoningWebSummerSchool} as well as articles
targeting a more general audience \citep{AustrianComputerSociety}.
More recently, this line of research ventured into other types of
ontological arguments, and a computer-assisted reconstruction of an
argument by Leibniz will appear as a chapter in a book dedicated to
the 300th anniversary of Leibniz's death
\citep{LeibnizOntologicalProof}. The automatic verification of G\"odel's
and Scott's variants \citep{ECAI2014} was preceded by an automatic
verification of (a non-modal version of) Anselm's ontological argument
\citep{ZaltaOppenheimer}. However, whereas the complexity of G\"odel's
and Scott's arguments required the embedding of higher-order modal
logics into higher-order logics and the use of higher-order automated
reasoners, Anselm's argument greater simplicity enabled it to be
verified with a first-order prover. Anselm's argument has also been
formalized within the interactive proof assistant PVS \citep{Rushby}
and there has been a so far failed attempt to formalize Avicenna's
argument in PVS as well.

@@

\begin{small}
\begin{prooftree}

@@

					\AXC{$ $} \RightLabel{5}
					\UIC{$G(\gamma)$} \dottedLine
					\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
					\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$} \RightLabel{$\imp_E$}
				\BIC{$\neg \rho(\gamma)$} \RightLabel{$\neg_E$}      
	   \BIC{$\bot$} \RightLabel{$\neg_I^1$}
	   \UIC{$\neg\neg P(\rho)$} \RightLabel{$\neg\neg_E$}
	   \UIC{$[\subproof_4] \qquad P(\rho)$}
	   \end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$G(\gamma)$}

 \AXC{$\subproof_4$}\dashedLine
 \UIC{$P(\rho)$}
	      \AXC{Axiom 4} \dashedLine
        \UIC{$\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]$} \RightLabel{$\all_E$}
	   	  \UIC{$P(\rho) \imp \nec P(\rho)$} \RightLabel{$\imp_E$}
		 \BIC{$\nec P(\rho)\scott$}
			\AXC{$ $} \RightLabel{3}
			\UIC{$\omega: P(\rho)$}
					\AXC{$ $} \RightLabel{2}
					\UIC{$\omega: G(\delta) $} \dottedLine
					\UIC{$\omega: \all \varphi. P(\varphi) \imp \varphi(\delta)$} \RightLabel{$\all_E$} 
					\UIC{$\omega: P(\rho) \imp \rho(\delta)$} \RightLabel{$\imp_E$}
				\BIC{$\omega: \rho(\delta) $} \RightLabel{$\imp_I^2$}
				\UIC{$\omega: G(\delta) \imp \rho(\delta) $} \RightLabel{$\all_I$}
				\UIC{$\omega: \all y. G(y) \imp \rho(y) $} \RightLabel{$\imp_I^3$}
				\UIC{$\omega: P(\rho) \imp (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\nec_I$}
				\UIC{$\nec (P(\rho) \imp (\all y. G(y) \imp \rho(y) ))$} \doubleLine \RightLabel{Axiom K}

					\UIC{$\nec P(\rho) \imp \nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\imp_E$}
			\BIC{$\nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\imp_I^4$}
			\UIC{$\rho(\gamma) \imp \nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\all_I$}
			\UIC{$\all \varphi. \varphi(\gamma) \imp \nec (\all y. G(y) \imp \varphi(y) )$} \RightLabel{$\wedge_I$}
		\BIC{$ G(\gamma) \wedge (\all \varphi. \varphi(\gamma) \imp \nec (\all y. G(y) \imp \varphi(y) ))$} \dottedLine
		\UIC{$\ess{G}{\gamma}$} \RightLabel{$ \imp_I^5$}
		\UIC{$G(\gamma) \imp \ess{G}{\gamma}$} \RightLabel{$\all_I$}
		\UIC{$\all x. G(x) \imp \ess{G}{x}$}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\neg P(\rho)$} 
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. [P(\neg \varphi) \biimp \neg P(\varphi)]$} \RightLabel{$\all_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \biimp \neg P(\rho)$} \RightLabel{$\wedge_E$}
\UIC{$\neg P(\rho) \imp P(\lambda x. \neg \rho(x))$}\RightLabel{$\imp_E$}
\BIC{$P(\lambda x. \neg \rho(x))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\forall \varphi.(P(\varphi) \imp \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\neg \rho(\gamma)$}
\AXC{$ $} \RightLabel{3}
\UIC{$\rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$\bot$}\RightLabel{$\imp_I^2$}
\UIC{$\neg\neg P(\rho)$}\RightLabel{$\neg\neg_E$}
\UIC{$[\subproof_7] \qquad P(\rho)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec P(\rho)$} \RightLabel{$\nec_E$}
\UIC{$\omega: P(\rho)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\omega: G(y)$}  \dottedLine \RightLabel{D\ref{N:D1}}
\UIC{$\omega: \forall \varphi.(P(\varphi) \imp \varphi(y))$}\RightLabel{$\forall_E$}
\UIC{$\omega: P(\rho) \imp \rho(y)$}\RightLabel{$\imp_E$}
\BIC{$\omega: \rho(y)$} \RightLabel{$\imp_I^4$}
\UIC{$\omega: G(y) \imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega: \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y.(G(y) \imp \rho(y))$} \RightLabel{$\imp_I^5$}
\UIC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$}

%\AXC{$ $} \RightLabel{6}
%\UIC{$\rho(\gamma)$} \RightLabel{}
\AXC{$ \subproof_7$} \dashedLine
\UIC{$P(\rho)$} %\RightLabel{$\nec_E$}
\AXC{Axiom \ref{N:A4}} \dashedLine
\UIC{$\all \varphi.[P(\varphi) \imp \nec\; P(\varphi)]$}\RightLabel{$\forall_E$}
\UIC{$P(\rho) \imp \nec\; P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$}\RightLabel{$\imp_E$}%\RightLabel{$\imp_I^3$}
%\UIC{$[\subproof_7] \qquad \rho(\gamma) \imp \nec\; P(\rho)$}
%\UIC{$\rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
%\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}

\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}  \RightLabel{$\wedge_I$}
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{N:D2}}
\UIC{$[\subproof_8] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{small}

@@

\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$} \RightLabel{}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\nec\rho(\gamma)$} \RightLabel{$\biimp_E$}
\BIC{$P(\rho)$} \RightLabel{$\forall_E$}

@@

\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$\nec P(\rho)$} \RightLabel{$\nec_E$}
\UIC{$\omega: P(\rho)$}
\AXC{$ $} \RightLabel{2}
\UIC{$\omega: G(y)$}  \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\omega: \forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$\omega: P(\rho) \biimp \nec \rho(y)$}\RightLabel{$\biimp_E$}
\BIC{$\omega: \nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega: \rho(y)$} \RightLabel{$\imp_I^2$}
\UIC{$\omega: G(y) \imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega: \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y.(G(y) \imp \rho(y))$} \RightLabel{$\imp_I^3$}
\UIC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$}

\AXC{$\subproof_7$} \dashedLine
\UIC{$\nec P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^4$}
\UIC{$[\subproof_8] \qquad \nec\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\end{prooftree}

%implication for equivalence in essence: 
%\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} \dashedLine
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_6$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. [P(\varphi)\biimp \nec \varphi(\gamma)]$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^5$}
\UIC{$[\subproof_9] \quad \nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
%\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$ and $\subproof_9$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $} \RightLabel{$\forall_I$}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)]) $}
\AXC{$ \subproof_6 $} \dashedLine
\UIC{$G(\gamma)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\biimp_E$}
\BIC{$\nec E(\gamma)$}\RightLabel{$\nec_E$}
   % \AXC{Axiom T} \dashedLine
    %\UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\UIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
%\end{tiny}
\end{proof}

@@

%\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(G) \biimp \nec G(\gamma)$}
\AXC{Axiom \ref{A:A3}} \RightLabel{}
\UIC{$P(G)$} \RightLabel{$\biimp_E$}
\BIC{$\nec G(\gamma)$}
\AXC{$ $} \RightLabel{2}
\UIC{$G(\gamma)$} \RightLabel{$\ex_I$}
\UIC{$\ex x.G(x)$}  \RightLabel{$\imp_I^2$}
\UIC{$G(\gamma)\imp \ex x.G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec [G(\gamma)\imp \ex x.G(x)]$} \doubleLine  \RightLabel{(Axiom K)}
\UIC{$\nec G(\gamma)\imp \nec \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\nec \ex x.G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z)\imp \nec \ex x.G(x)$}
\end{prooftree}
%\end{scriptsize}
